---
title: "Modeling"
author: "STA310"
date: "2023-11-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)    # graphics package
library(dplyr)      # joining data frames and data manipulation
library(leaps)
library(tidyr)
library(corrplot)
library(mosaic)
library(patchwork)
```

## Read in Files
```{r}
Merged3 <- read.csv("merged/Monthly.Merged.csv")
Y.Merged2 <- read.csv("merged/Yearly.Merged.csv")
```

### 1. Modeling Approach (Monthly)

Before checking best subset, we will delete redundant columns and Month (since all Months appears to be important predictor).
```{r}
# Delete extra columns (without Month)
Merged4 <- select(Merged3, -c(NCESDistrictID, year, entity_name, district_name,
                              report_school_year, pop.5.17, num.report.school,
                              NewCasesTeachers, NewCasesStaff, AccuCasesStudents, AccuCasesTeachers,
                              AccuCasesStaff, Norm.New.Stud, share_virtual, num_counselors,
                              num_social, num_princ, District.Name, Month))


# Delete extra columns (keep Month)
Merged5 <- select(Merged3, -c(NCESDistrictID, year, entity_name, district_name,
                              report_school_year, pop.5.17, num.report.school,
                              NewCasesTeachers, NewCasesStaff, AccuCasesStudents, AccuCasesTeachers,
                              AccuCasesStaff, Norm.New.Stud, share_virtual, num_counselors,
                              num_social, num_princ, District.Name))

Merged6 <- na.omit(Merged5)
```


```{r}
# Best subset
bestlm <- regsubsets(LaggedStudCases ~ ., data = Merged4, method="exhaustive")
plot(bestlm, scale = "adjr2")
summary(bestlm)$adjr2

# Check Correlations Plot
correlations <- cor(Merged4, use = "pairwise.complete.obs")
corrplot(correlations)
```

Based on best subset result, we add the important predictor in our main effects model.
```{r}
## model0 : Main effect Model
model0 <- lm(log(LaggedStudCases+1) ~ Month + share_inperson + NewCasesStudents + 
               per_free_lunch + pupil_count_tot + Teacher_Sal,
             data = Merged6)
summary(model0)
```

Then, we create the interaction model to identity important interaction terms needed to add into our model. We check R-square value to make sure the predictive power of our model.
```{r}
## model1 : Check all interaction
model1.m <- lm(log(LaggedStudCases+1) ~ (Month + share_inperson + NewCasesStudents + 
                                         per_free_lunch + pupil_count_tot + Teacher_Sal)^2,
             data = Merged6)
summary(model1.m)
```

Our final monthly model includes only interaction terms that has observable effects on R-square.
```{r}
## model2 : Interaction Model
model2.m <- lm(log(LaggedStudCases+1) ~ Month + share_inperson + NewCasesStudents + 
               per_free_lunch + pupil_count_tot + Teacher_Sal +
               pupil_count_tot*Teacher_Sal + per_free_lunch*pupil_count_tot + 
               NewCasesStudents*per_free_lunch + NewCasesStudents*pupil_count_tot +
               Month*per_free_lunch + Month*pupil_count_tot,
             data = Merged6)
summary(model2.m)

# Check Residual Plot
mod.res <- model2.m$residuals
qqnorm(mod.res)
```

### 2. Modeling Approach (Yearly)

(Modeling for yearly data follows similar approach as monthly)
```{r}
Y.Merged3 <- Y.Merged2 %>%
  select(-c(year, entity_name, district_name,
            report_school_year, NewCasesTeachers, NewCasesStaff, Norm.New.Stud, 
            share_virtual, num_counselors, num_social, num_princ)) %>%
  rename("TotalCasesStudent" = "NewCasesStudents")
Y.Merged4 <- na.omit(Y.Merged3)
```

```{r}
model0.y <- lm(TotalCasesStudent ~ Teacher_Sal + student_teacher_ratio + 
                 per_free_lunch + math_per_prof.18  + pupil_count_tot +  share_inperson,
               data = Y.Merged4)
summary(model0.y)


# Check Correlations Plot
correlations <- cor(Y.Merged4, use = "pairwise.complete.obs")
corrplot(correlations)

## model1 : Check all interaction 
model1.y <- lm(TotalCasesStudent ~ (Teacher_Sal + student_teacher_ratio + 
                                      per_free_lunch + math_per_prof.18  + pupil_count_tot + 
                                      share_inperson)^2,
               data = Y.Merged4)
summary(model1.y)


## model2 : Interaction Model
model2.y <- lm(log(TotalCasesStudent+1) ~ Teacher_Sal*Teacher_Sal + Teacher_Sal + 
                 student_teacher_ratio + per_free_lunch + math_per_prof.18  + pupil_count_tot + 
                 share_inperson + pupil_count_tot*share_inperson + per_free_lunch*pupil_count_tot +
                 student_teacher_ratio*pupil_count_tot,
               data = Y.Merged4)

summary(model2.y)
```


### 3. Plotting

Create categories for continuous variables (for side-by-side plot).
```{r}
Y.Merged4$NormalizedCases = 1000 * (Y.Merged4$TotalCasesStudent / Y.Merged4$pupil_count_tot)
Y.Merged4$CasesCategory <- cut(Y.Merged4$NormalizedCases, 
                               breaks = c(0, 30, 60, 90, Inf),
                               labels = c("0-30", "30-60", "60-90", ">90"),
                               include.lowest = TRUE)

Y.Merged4$SalaryCategory <- cut(Y.Merged4$Teacher_Sal, 
                                breaks = c(50003, 62340, 69744, 95366, Inf),
                                labels = c("Low Salary (<25)", "Below Median Salary", "Above Median Salary", "High Salary (>75)"),
                                include.lowest = TRUE)
```

```{r}
# Histogram for TotalCasesStudent
p1 <- ggplot(Y.Merged4, aes(x = TotalCasesStudent)) +
  geom_histogram(aes(y = ..count..), bins = 30, fill = "#1f77b4", alpha = 0.6) +
  geom_density(aes(y = 33*..count..), color = "blue", size = 1) +
  labs(title = "Histogram of TotalCasesStudent") +
  theme_minimal()
p1

# Histogram for NormalizedCases
p2 <- ggplot(Y.Merged4, aes(x = NormalizedCases)) +
  geom_histogram(aes(y = ..count..), bins = 30, fill = "#1f77b4", alpha = 0.6) +
  geom_density(aes(y = 4.5*..count..), color = "blue", size = 1) +
  labs(title = "Histogram of NormalizedCases (per 1000 pupils)") +
  theme_minimal()
p2

# Histogram for share_inperson
p3 <- ggplot(Y.Merged4, aes(x = per_free_lunch)) +
  geom_histogram(aes(y = ..count..), bins = 30, fill = "#9467bd", alpha = 0.6) +
  geom_density(aes(y = 3.2*..count..), color = "purple", size = 1) +
  labs(title = "Histogram of per_free_lunch") +
  theme_minimal()
p3

# Histogram for pupil_count_tot
p4 <- ggplot(Y.Merged4, aes(x = pupil_count_tot)) +
  geom_histogram(aes(y = ..count..), bins = 30, fill = "#9467bd", alpha = 0.6) +
  geom_density(aes(y = 1000*..count..), color = "purple", size = 1) +
  labs(title = "Histogram of pupil_count_tot") +
  theme_minimal()
p4

# Histogram for Teacher_Sal
p5 <- ggplot(Y.Merged4, aes(x = Teacher_Sal)) +
  geom_histogram(aes(y = ..count..), bins = 30, fill = "#9467bd", alpha = 0.6) +
  geom_density(aes(y = 2850*..count..), color = "purple", size = 1) +
  labs(title = "Histogram of Teacher_Sal") +
  theme_minimal()
p5

# Histogram for student_teacher_ratio
p6 <- ggplot(Y.Merged4, aes(x = student_teacher_ratio)) +
  geom_histogram(aes(y = ..count..), bins = 30, fill = "#9467bd", alpha = 0.6) +
  geom_density(aes(y = 0.555*..count..), color = "purple", size = 1) +
  labs(title = "Histogram of student_teacher_ratio") +
  theme_minimal()
p6

# Arranging plots and Print out
layout1 <- (p1 | p2)
layout2 <-  (p3 | p4) /
  (p5 | p6)
#layout1
#layout2
```

Based on the bimodel shape of teacher salary, we create the following plot facetted by different levels of teacher salary.
```{r}
# % of Free Lunch vs. (log) Student Cases by Teacher Salary
ggplot(Y.Merged4, aes(x = per_free_lunch, y = log(TotalCasesStudent))) +
  geom_point(color = "#1f77b4") +  # Add color to the points
  facet_wrap(~SalaryCategory) +
  labs(title = "% of Free Lunch vs. (log) Student Cases by Teacher Salary",
       x = "Percentage of Free Lunch",
       y = "(log) Total Cases") +
  theme_minimal()
```


